---
title: "test-mcda"
author: "Emily Zhou"
date: "2024-05-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Testing The MCDA Package

```{r}
Rglpk, glpkAPI, RColorBrewer, combinat, triangle, plyr,
ggplot2
```


## Testing the AHP Approach 
```{r}
alts <- c("Corsa","Clio","Fiesta","Sandero")


style <- matrix(c(1.0, 1/4, 4.0, 1/6,
4.0, 1.0, 4.0, 1/4,
1/4, 1/4, 1.0, 1/5,
6.0, 4.0, 5.0, 1.0),
nrow=length(alts), ncol=length(alts), byrow=TRUE,
dimnames=list(alts,alts))


reliability <- matrix(c(1.0, 2.0, 5.0, 1.0,
1/2, 1.0, 3.0, 2.0,
1/5, 1/3, 1.0, 1/4,
1.0, 1/2, 4.0, 1.0),
nrow=length(alts), ncol=length(alts), byrow=TRUE,
dimnames=list(alts,alts))


fuel <- matrix(c(1.0, 2.0, 4.0, 1.0,
0.5, 1.0, 3.0, 2.0,
1/4, 1/3, 1.0, 1/5,
1.0, 1/2, 5.0, 1.0),
nrow=length(alts), ncol=length(alts), byrow=TRUE,
dimnames=list(alts,alts))
```


```{r}

alternativesPairwiseComparisonsList <- list(style = style,
reliability = reliability,
fuel = fuel)

```


```{r}
crit <- c("style","reliability","fuel")
criteriaWeightsPairwiseComparisons <- matrix(c(1.0, 1/2, 3.0,
2.0, 1.0, 4.0,
1/3, 1/4, 1.0),
nrow=length(crit),
ncol=length(crit),
dimnames=list(crit,crit))
```


```{r}
AHP(criteriaWeightsPairwiseComparisons, alternativesPairwiseComparisonsList)
```

## Testing the TOPSIS Approach

```{r}


performanceTable <- matrix(c(5490,51.4,8.5,285,6500,70.6,7,
                             288,6489,54.3,7.5,290),
                           nrow=3,
                           ncol=4,
                           byrow=TRUE)

row.names(performanceTable) <- c("Corsa","Clio","Fiesta")
colnames(performanceTable) <- c("Purchase Price","Economy",
                                "Aesthetics","Boot Capacity")


weights <- c(0.35,0.25,0.25,0.15)
criteriaMinMax <- c("min", "max", "max", "max")

# positiveIdealSolutions <- c(0.179573776, 0.171636015, 0.159499658, 0.087302767)
# negativeIdealSolutions <- c(0.212610118, 0.124958799, 0.131352659, 0.085797547)
names(weights) <- colnames(performanceTable)

names(criteriaMinMax) <- colnames(performanceTable)
# names(positiveIdealSolutions) <- colnames(performanceTable)
# names(negativeIdealSolutions) <- colnames(performanceTable)
```


```{r}
overall1 <- TOPSIS(performanceTable, weights, criteriaMinMax)
overall1 # returns a vector containing the TOPSIS score for each alternative
```

## Testing the PROMETHEE Approach
Uses preference functions and outranking flows to rank alternatives based on pairwise comparisons.

1. Calculate the difference in performance scores for each criterion between all pairs of alternatives.
2. Set up an indifference threshold.
3. Set up a preference threshold. 
4. Decide a preference function
5. Apply the preference function using the indifference and preference threshold to convert the differences into preference value, which is typically betweeen 0 and 1. 
6. For each alternative, sum the preference values across all criteria to obtain the aggregated preference indices.
7. Calculate the positive and negative outranking flows for each alternative, which represent how much an alternative outranks others and is outranked by others, respectively.
8. Use the net outranking flow (positive flow minus negative flow) to rank the alternatives.


The gaussParameter vector in the code specifies the standard deviation (s) for the Gaussian preference function for each criterion.

Terminology: 
- preference function: define how much better one site is better over another for each criterion (a linear function where the lower the accessibility, the better)

- preference threshold: the minimum difference between the performance of two alternatives on a given criterion such that the decision-maker has a strict preference for one alternative over the other. **If the performance difference is greater than or equal to p, the decision-maker clearly prefers one alternative over the other for that criterion. **

- indifference threshold: The indifference threshold q represents the maximum difference between the performance of two alternatives on a given criterion such that the decision-maker is indifferent between them. **If the performance difference is less than or equal to q, the decision-maker considers the two alternatives to be effectively equal with respect to that criterion.**


```{r}

performanceTable <- rbind(
c(1,10,1),
c(4,20,2),
c(2,20,0),
c(6,40,0),
c(30,30,3))
rownames(performanceTable) <- c("RER","METRO1","METRO2","BUS","TAXI")
colnames(performanceTable) <- c("Price","Time","Comfort")
```

```{r}
# The preference functions
preferenceFunction<-c("Gaussian","Level","V-shape-Indiff")
#Parameter of the Gaussian preference function
gaussParameter<-c(4,1,0)
names(gaussParameter)<-colnames(performanceTable)
```

```{r}
#Preference threshold
preferenceThreshold<-c(5,15,3)
names(preferenceThreshold)<-colnames(performanceTable)
```

```{r}
#Indifference threshold
indifferenceThreshold<-c(3,11,1)
names(indifferenceThreshold)<-colnames(performanceTable)
```


```{r}
#weights
criteriaWeights<-c(0.2,0.3,0.5)
names(criteriaWeights)<-colnames(performanceTable)

# criteria to minimize or maximize
criteriaMinMax<-c("min","min","max")
names(criteriaMinMax)<-colnames(performanceTable)
```


```{r}
PROMETHEEOutrankingFlows(performanceTable, preferenceFunction,preferenceThreshold,
indifferenceThreshold,gaussParameter,criteriaWeights,criteriaMinMax)
```






## Testing the AHP Approach


